---
title: "Faculty Classification"
author: Chad Evans
output: 
  github_document:
  toc: true
always_allow_html: yes
params:
 d: !r Sys.Date() 

---
Built with `r getRversion()`.  Last run on `r params$d`.

* [Configure](#config)
    + Libraries
    + directories
    + data
* [Munge](#munge)
* [Exploratory Analysis](#exploratory)
    + [Missing Data](#missing)
    + [Single Imputation](#impute)
    

## Configure{#config}
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.width=10, fig.height=10, fig.path='graphs/', cache.path ='cache/')
```

```{r Libraries, include=FALSE, eval=TRUE, echo=TRUE, warning=TRUE,error=FALSE, message=TRUE, tidy=TRUE, results='markup', cache=FALSE, fig.width=7, fig.height=7}
library(tidyverse)
library(stats)
library(mice)
library(knitr)
```

```{r Directories, include=FALSE}
Private_Cache <- "/Users/chadgevans/Research/Projects/Data/HERI/Cache"
Raw<-"/Users/chadgevans/Research/Projects/Data/HERI/Raw"
Munge<-"./munge"
Source<-"./src"
Graph<-"./graphs"
```

## Munge{#munge}
```{r Data Munge, eval=FALSE, cache=TRUE}
source(file.path(Munge, "01_Merge_the_data.R"))
source(file.path(Munge, "02_Clean_the_data.R"))
source(file.path(Munge, "Recode_HERI.R"))
save(df, file=file.path(Private_Cache,"HERI_Class.RData"))
```

```{r Clean Data Import}
load(file.path(Private_Cache,"HERI_Class.RData"))
```


 Multiple Imputation was the preferred method when our
data structure was more complicated (high dimensions, high cluster overlap),

```{r}
# Present Day Work Conditions (also opinions and attitutdes)
#Classify faculty based on job attitudes/activities/experiences
SVYVARS<-c("ACE","SUBJID","YEAR","ACERECODE","RESTYPE")
INSTVARS<-c("DEPTA","INSTTYPE","INSTCONT","SELECTIVITY","CARNEGIE")
DEMGVARS<-c("GENACT03","GENACT04","NCHILD1","NCHILD2","MARITAL","SEX","NATENGSP","RACEGROUP","STATE","OBEREG","MAJORA","AGE")
PTVARS<-c("PTCHOICE","PTWORKFT","PTTSEEK","PTCAREER","PTREASON01","PTREASON02","PTREASON03","PTREASON04","PTREASON05","PTREASON06","PTREASON07","PTRESOURCES01","PTRESOURCES02","PTRESOURCES03","PTRESOURCES04","PTRESOURCES05","PTOPN01","PTOPN02","PTOPN03","PTOPN04","PTOPN05","PTOPN06","PTOPN07","PTOPN08","PTOPN09","PTOPN10","PTTEACH","PTSALARY","PTPAY")
OTHERS<-c("DEPTDISC","SSVALID","SALARY","TENUREYR","SALARY12","CCRANK","CCSTATUS","SALARYBASE","BIRTHYR")

#HIGHMISSVARS<-c()
OMIT<-c(SVYVARS,INSTVARS,DEMGVARS,PTVARS,OTHERS)
```

## Exploratory Analysis{#explore}

# Missing Data{#missing}

```{r}
miss_pct<-df %>% 
  select(-one_of(OMIT)) %>%
  map_dbl(function(x) { round((sum(is.na(x)) / length(x)) * 100, 1) })
data.frame(miss=miss_pct, var=names(miss_pct), row.names=NULL) %>%
ggplot(aes(x=reorder(var, -miss), y=miss)) +
geom_bar(stat='identity', fill='red') +
labs(x='', y='% missing', title='Percent missing data by feature') +
theme(axis.text.x=element_text(angle=90, hjust=1))
```

## Single Imputation{#impute}

```{r Impute_data, cache=TRUE}
tempData<-df %>% 
  select(-one_of(OMIT)) %>%
  mice(m=1,maxit=50,meth='pmm',seed=500)
completedData <- complete(tempData,1)
```

```{r Create_Binaries}
dfClust<-data.frame(model.matrix(~ ., data=completedData, contrasts.arg = lapply(completedData[,sapply(completedData, is.factor)], contrasts, contrasts=FALSE)))
```

```{r Determin_Number_Clusters}
wssplot <- function(data, nc=15, seed=1234){
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")}
wssplot(dfClust, nc=7) # put in # of clusters here
```

```{r K_means}
Clust<- dfClust %>%
  kmeans(5, nstart = 20)
Clust
```

```{r add_clusters}
CLUSTER<-Clust$cluster
df<-cbind(df,CLUSTER)
```


```{r demography_table}
options(digits=2)
list<-do.call(rbind, sapply(df[DEMGVARS], function(x){
    if(is.numeric(x)==TRUE){
      means<-aggregate(x ~ df$CLUSTER, FUN=mean, na.action=na.omit)
      print(means)[,2]
      }
    else{
      prop.table(table(x, df$CLUSTER),2)
      }
  }))
kable(list, caption = "Distribution of Adjunct Types by Demographic Characteristics")
```


```{r Institution_table}
list<-do.call(rbind, sapply(df3[INSTVARS], function(x){
    if(is.numeric(x)==TRUE){
      means<-aggregate(x ~ df$CLUSTER, FUN=mean, na.action=na.omit)
      print(means)[,2]
      }
    else{
      prop.table(table(x, df$CLUSTER),2)
      }
  }))
kable(list, caption = "Distribution of Adjunct Types by Institutional Characteristics")
```


